\chapwithtoc{Conclusion}

This thesis summarizes several efforts to improve the performance of complex algorithms with inefficient implementations from contemporary scientific domains, improving their practical usability by applying high-performance computing principles. In the two chapters, it outlines the motivations and results of six research contributions, presented in \cref{chap:contributions}.

The first four contributions detail the parallelization and optimization challenges of the scientific algorithms which we have worked on during our research. The results include GPU applications that use novel data structures, promote high scalability, and utilize complicated GPU optimization techniques. The presented implementations provide orders of magnitude speedups over the prior state-of-the-art, enabling domain scientists to finish their analyses in minutes instead of days, process more data with greater accuracy, interactively visualize the results, and explore previously unattainable problem variations. The methodology of the works can serve as a helpful support for other researchers in the domain of GPGPU computing.

The last two contributions present the use cases of the novel HPC library Noarr, specializing in expressing the layout and traversal of $n$-dimensional arrays, which are the most commonly used data structures in scientific computing. The novel approach of assigning names to the dimensions of the arrays and expressing the layout and traversal in a declarative way allows users to deploy memory-related optimizations in a more readable and maintainable manner while the library takes care of the complex indexing and loop transformations.

As implied throughout the thesis, the use of GPUs for general computing is becoming increasingly popular. Consequently, these devices improve in versatility with each addition of new core types, specialized high-throughput instructions, and new thread hierarchies. This creates unique and interesting opportunities for future research: The implementation of Mahalanobis Hierarchical Clustering can be directly extended with the use of tensor cores, which may provide an additional order of magnitude performance improvement thanks to their high compute bandwidth. Similarly, we plan to continue on $k$-NN development, which we have already started during the work on EmbedSOM, as we believe that increasing the caching capabilities by employing the new feature of distributed shared memory may improve the data throughput of such a highly memory-bound algorithm. To support the efforts, we plan to continue extending the Noarr library towards the auto-tuning direction, ultimately providing a machine learning-guided optimizer of the array layouts and traversals. We also plan to finish our work on biological simulations, especially BioFVM and PhysiCell, which has already served as a valuable case study for Noarr and has partly guided its development thanks to the presence of multiple complex high-dimensional arrays, which accelerated its usability.

Finally, we hope that the contributions of this thesis will help to develop software that performs well in spite of the impending technical challenges and progressive widening of the bandwidth-compute gap. We believe that the presented methodologies and contributions provide valuable insights that can serve as a foundation for further research in the field of GPGPU computing and HPC.
